Non-word repetition in Yélî Dnye

### 2020-12-21 manuscript.Rmd
add indiv diffs disc subsection draft

Previous work typically steers clear of articulartorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology. We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs. less challenging sounds. However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below). Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items. However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit. That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.

Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech], and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

@@

- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
- How does score change as a function of item length in number of syllables?
- Is individual variation in NWR scores attributable to child age, sex, birth order, monolingual status, and/or parental education?

@@

We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item and its average segmental frequency. Based on previous work, we had predicted that children would have higher NWR scores for shorter items. That said, previous work has shown both very small [@piazzalunga2019articulatory] and very large [@cristia2020infant; @jabere2018xperiment] effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10% of the remaining words. Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms **Alex I'm going off your note here ("Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.""). I don't have a reference for this, can you please finish the thought or nix this bit?**. Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long. While indeed NWR scores were overall lower for longer items (e.g., see Figure \@ref(fig:prevlit-fig)), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant. In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length. First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon. Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s [-@piazzalunga2019articulatory] study on Italian and as borne out in the current dataset once controlling for other factors.

@@

- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?

3. Children's scores increases with child age.

5.  General prediction: previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated score higher in standardized language tests (eg ppvt) than children whose mothers are less educated. 
5. Prediction made before seeing data: In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment. This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly. However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
5. We did not find this

6. General prediction: To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children score higher in standardized language tests later-born children. 
6. Prediction before seeing data: One main causal path between birth order and language development is via parental input (CITE). Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the difference in scores between first borns and later borns may be small
6. we did not find a sig effect, but this is a small effect (eg d=.2 in Havron et al)

>>>>

Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology. We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs. less challenging sounds. However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below). Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items. However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit. That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.

Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]<!--, and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]-->. Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool]. Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.

@@

- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
- How does score change as a function of item length in number of syllables?
- Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?

@@

We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item and its average segmental frequency. Based on previous work, we had predicted that children would have higher NWR scores for shorter items. That said, previous work has shown both very small [@piazzalunga2019articulatory] and very large [@cristia2020infant; @jabere2018xperiment] effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10% of the remaining words. Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms **M2A: Alex I'm going off your note here ("Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.""). I don't have a reference for this, can you please finish the thought or nix this bit?**. Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long. While indeed NWR scores were overall lower for longer items (e.g., see Figure \@ref(fig:prevlit-fig)), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant. In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length. First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon. Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s [-@piazzalunga2019articulatory] study on Italian and as borne out in the current dataset once controlling for other factors.

@@

A review of previous work (see Introduction) suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR. We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta-analytic efforts aggregating over smaller studies such as ours.

Following prior work, we expected that NWR scores would increase with participant age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency. In brief, our results underscore the idea that phoonlogical development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].

In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or participant gender [@chiat2007preschool] on NWR scores. In addition to this prior work, education on Rossel Island, while generally highly valued, is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation. There is also limited variation in maternal education across the families in the region of the island where we sampled. We therefore expected little evidence for impact of either participant gender or maternal education in the present study. On the other hand, these predictors have established effects on other language development measures (**REFS: M2A: Alex go ahead and pick your faves here**). So to the extent that NWR scores share causal links to gender-based differences ein development and maternal linguistic input with these other language ouctome measures, we might then expect these factors to appear in NWR data. In fact, neither participant gender nor maternal education were correlated with NWR score in the current data.

Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect], presumably because later-born children receive a smaller share of maternal input than their older siblings. Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers. Therefore we expected that any effects of birth order on NWR would be attenuated in this context. In line with this prediction, our descriptive analysis showed no correlation between birth order and NWR score.
